---
title: "storm-report"
author: "Ed Kademan"
date: "December 20, 2015"
output: html_document
---

## Synopsis

This report summarizes the economic and health consequences of storms
and other extreme weather events. The economic data for years prior
to 2000 are very sparse so I just reported the aggregate dollar
amount of property and crop damage by event for the entire dataset.
Floods, hurricanes, tornados and ocean-related events such as storm
surges and erosion were the most costly over that period.

Tornados, other wind storms, heat and flood caused the most number
of injuries and fatalities. There were many injury reports from
early (20th century) years but they mainly involved tornados.
I suspect that the numbers of deaths and so on due to non-tornado
events simply didn't get reported.

## Data Processing

I downloaded the file that this
[link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 "csv file")
points to. Because it is so large and time-consuming to read I cache the results when I load it into R.

```{r cache = TRUE}
storm_data <- read.csv("repdata-data-StormData.csv.bz2")
```

The code below converts the storm data into a form that is easier
to work with. It discards any records for which there is neither
injury/fatality information nor readable damage estimates. Then it
converts the damage estimates to a straightforward dollar amount.
It also regularizes the events described. There were many `EVTYPE`
values that belonged together but that differed from on another in
very insignificant ways. For example, `SNOW SQUALL` and
`SNOW SQUALLS`.

```{r warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)

low <- function(x) casefold(as.character(x))

## This consolidates the event descriptions and fixes spelling
## mistakes. It reduces the number of unique events from 483 to
## 61.
simplify_events <- function(events) {
  r <- as.character(events)
  g <- function(pat) grep(pat, r, ignore.case = TRUE)
  r[g("wind")] <- "wind"
  r[g("thunderstorm|tstm|lightning")] <- "thunderstorm"
  r[g("rain|shower|burst")] <- "rain"
  r[g("snow|wintry|winter|blizzard")] <- "snow"
  r[g("cold|freez|ice|sleet|glaze|low temper|icy|frost")] <- "cold"
  r[g("hurricane")] <- "hurricane"
  r[g("torn|funnel")] <- "tornado"
  r[g("hail")] <- "hail"
  r[g("flood")] <- "flood"
  r[g("tropical storm")] <- "tropical storm"
  r[g("heat|warm weather")] <- "heat"
  r[g("current|coast|seas|marine|surf|beach|wave|surge|swell")] <- "coastal/seas"
  r[g("fire")] <- "fire"
  r[g("slid|slump|avala")] <- "landslide"
  r[g("fog")] <- "fog"
  factor(low(r))}

noaa_date_to_decade <- function(x) {
  ## extract the 4-digit year
  x <- sub("\\d{1,2}/\\d{1,2}/(\\d{4}).*", "\\1", as.character(x))
  cut(as.numeric(x),
      breaks = seq(from = 1949, to = 2019, by = 10),
      labels = c("50s", "60s", "70s", "80s",
                 "90s", "00s", "10s"),
      ordered_result = TRUE)}

read_storm_data <- function(d = readRDS("storm-data.rds")) {
  validexp <- function(f) low(f) %in% c("k", "m", "b")
  scale_val <- function(v, s) {
    i <- v > 0
    v[i] <- v[i]*c(k = 1e3, m = 1e6, b = 1e9)[low(s[i])]
    v}
  d %>%
    filter(INJURIES > 0 | FATALITIES > 0 |
           (PROPDMG > 0 & validexp(PROPDMGEXP)) |
           (CROPDMG > 0 & validexp(CROPDMGEXP))) %>%
    mutate(PROPDMG = scale_val(PROPDMG, PROPDMGEXP),
           CROPDMG = scale_val(CROPDMG, CROPDMGEXP),
           Event   = simplify_events(EVTYPE),
           Decade  = noaa_date_to_decade(BGN_DATE))}

d <- read_storm_data(storm_data)
```

## Results

### Dollar Amounts

We will look at property damage and crop damage first. The plot
below summarizes those dollar amounts by event for the entire
6 decades the data cover.

```{r}
costly <- function(d = read_storm_data()) {
  r <- d %>%
    group_by(Event) %>%
    summarise(Total_Prop = sum(PROPDMG, na.rm = TRUE),
              Total_Crop = sum(CROPDMG, na.rm = TRUE)) %>%
    mutate(Total_Dollar = Total_Prop + Total_Crop) %>%
    arrange(desc(Total_Dollar))
  r$Cumu_Prop <- cumsum(r$Total_Dollar)/sum(r$Total_Dollar)
  r}

costly_plot <- function(d = read_storm_data(), prop = .999) {
  d <- within(filter(costly(d), Cumu_Prop <= prop),
              Event <- factor(Event, levels = rev(Event),
                               ordered = TRUE)) %>%
    filter(Cumu_Prop <= prop) %>%
    melt(measure.vars = c("Total_Prop", "Total_Crop")) %>%
    within(variable <- c(Total_Prop = "Property Damage",
                         Total_Crop = "Crop Damage")[variable])
  ggplot(d) +
    geom_bar(aes(Event, value*1e-9, fill = variable),
             stat = "identity", position = "dodge") +
    ylab("billions of dollars") +
    coord_flip()}

costly2 <- function(d = read_storm_data())
  d %>%
    group_by(Decade, Event) %>%
    summarise(Total_Prop = sum(PROPDMG, na.rm = TRUE),
              Total_Crop = sum(CROPDMG, na.rm = TRUE))

costly2_plot <- function(d = read_storm_data(), n = 10) {
  ordered_event <- costly(d)$Event
  d <- costly2(d) %>%
    mutate(Event = factor(low(Event), levels = rev(ordered_event),
                          ordered = TRUE)) %>%
    filter(Event %in% ordered_event[seq(from = 1, to = n)]) %>%
    melt(measure.vars = c("Total_Prop", "Total_Crop"))
  ggplot(d) +
    geom_bar(aes(Event, value*1e-9, fill = variable),
             stat = "identity", position = "dodge") +
    ylab("billions of dollars") +
    coord_flip() +
    facet_wrap(~ Decade)}

costly_plot(d)
```

These are only the most costly weather events. The remainder that
the plot leaves out accounted for less than 0.1% of the damage.

### Injuries and Deaths

Summarize the numbers of injuries and fatalities overall by event
and by event/decade.

```{r}
deadly <- function(d = read_storm_data(),
                   group_by_vars = ~Event) {
  d %>%
    filter(INJURIES > 0 | FATALITIES > 0) %>%
    group_by_(.dots = group_by_vars) %>%
    summarise(Injuries   = sum(INJURIES)/1000,
              Fatalities = sum(FATALITIES)/1000) %>%
    arrange(desc(Injuries + Fatalities))}

deadly_plot <- function(d = read_storm_data(), n = 20) {
  d <- within(
    deadly(d),
    Event <- factor(Event, levels = rev(Event), ordered = TRUE))
  d <- melt(d[seq(from = 1, to = n),], id.vars = "Event")
  ggplot(d) +
    geom_bar(aes(Event, value), stat = "identity") +
    facet_wrap(~ variable) + ylab("thousands") +
    coord_flip()}

deadly2_plot <- function(d = read_storm_data(), n = 10) {
  ordered_event <- deadly(d)$Event
  d <- deadly(d, group_by_vars = list(~Decade, ~Event)) %>%
    mutate(Event = factor(low(Event), levels = rev(ordered_event),
                          ordered = TRUE)) %>%
    filter(Event %in% ordered_event[seq(from = 1, to = n)]) %>%
    melt(measure.vars = c("Injuries", "Fatalities"))
  ggplot(d) +
    geom_bar(aes(Event, value, fill = variable),
             stat = "identity", position = "dodge") +
    ylab("thousands") +
    coord_flip() +
    facet_wrap(~ Decade)}
```

The plots that the above functions create describe the data as below.

The `r n <- 20; n` most severe events for the entire period:

```{r echo = FALSE, fig.scap = "Health Consequences from 1950 through 2011"}
deadly_plot(d, n = n)
```

and the `r n <- 10; n` most severe by decade:

```{r echo = FALSE}
deadly2_plot(d, n = n)
```

